Local Sensitivity Analysis of a Supercritical Extraction Model

### 2025-03-29 Article_Sensitivity_Analsysis-Chamomile.tex
29.03

@@

	%\subfile{Sections/Results_Sensitivity}
	\section{Results} \label{CH: Results}
	The details of the process model and parameters are provided in \citet{Sliczniuk2024}. The process model was validated in the following range of operating conditions: temperatures between $30 - 40 ^\circ C$, pressures between $100 - 200$ bar, and mass flow rates between $3.33-6.67 \cdot 10^{-5}$ kg/s. This study examines the impact of parameter change on the state space and extraction yield through local sensitivity analysis. The local sensitivity analysis results can be interpreted as the system's response to an infinitesimal deviation in parameters. {\color{blue}This work focus on two selected cases, but the same analysis can be extended to any other model parameter. The first case corresponds to the analysis of empirical correlations on the process model. The zero'th order term (denoted as $\Upsilon(0)$) of the multiple linear regression for $\Upsilon$ was selected. The second case corresponds to the investigation of the operating condition on the state-space. Particularly, on the influence of pressure variations on the model output. The first analysis is performed at the midpoint of the validated range: $35 ^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s. The second analysis is splitted into two sub-analysis. The influence of the pressure change on the state-variables (concnetration of the solute in the sold and fluid phases) are considered at the midpoint of the validated range. The yield analysis is conducted at the same tempature and mass flow-rate conditions but at different values of pressure (100, 125, 150, 175 and 200 bar).}

@@

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CS_gamma.png}
		\caption{The effect of $P_{in}$ change on $C_s$}
		\label{fig:Sensitivty_Gamma_CS}
	\end{figure}

	The $\Upsilon$ term corresponds to the curvature of decay function $\gamma$, and can be associated with uneven distribution of the solute in the solid particle. As the zero'th order term is not directly affect by independent variables, it physical interpretation becomes straight forward. The increment of the term makes the exponential decay more steep, which means that the $D_i^R$ is multiplied by a smaller number, hence it is expected to observe slower extraction rate.

	Figure \ref{fig:Sensitivty_P_CS} illustrates the sensitivity of the solute concentration in the solid phase to change into zero'th order term in the equation for $\Upsilon$. As discussed by \citet{Sliczniuk2024}, the system is far from the saturation point, which indicates that the concentration gradient is controlled mainly by $c_s$. Considering that, the extraction kinetic term act as independent on $c_f$, which explain why Figure \ref{fig:Sensitivty_P_CS} shows a uniform response in solid phase. The positive value of $\frac{dc_s}{d\Upsilon(0)}$ can be interpreted as an slower loss of the solute from particles and is observed when the extraction slows down.

@@

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CS_P.png}
		\caption{The effect of $P_{in}$ change on $C_s$}
		\label{fig:Sensitivty_P_CS}
	\end{figure}

@@

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CF_P.png}
		\caption{The effect of $P_{in}$ change on $C_f$}
		\label{fig:Sensitivty_P_CF}
	\end{figure}

@@

	One disadvantage of using local sensitivity coefficients is that they reflect absolute changes, which can be challenging to compare across different models or scenarios. To address this, normalized local sensitivity coefficients (defined as: $\tilde{S}(x;\Theta) = \frac{d G(x;\Theta)}{d\Theta} \frac{\Theta}{G(x;\Theta)}$) are used. They can be interpreted as the percentage change in the output per percentage change in the parameter. Their dimensionless nature makes the invariant under rescaling of model parameters.

@@

	Sensitivity analysis is a tool used to understand how the parameters of the model affect its output. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. This work implemented automatic differentiation to derive the sensitivity equations. The results of the local sensitivity analysis and model specific and consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions.

	Although local sensitivity analysis can be performed with respect to any model parameter, this study focuses specifically on the effect of pressure. Under the selected operating conditions, it was observed for all the cases that a pressure increment enhances mass transfer, leading to a faster mass transfer of solute from the particles. Consequently, negative sensitivities can be observed in the diagram for the solid phase. The corresponding response in the fluid phase is characterized by positive-valued sensitivities, indicating that a greater amount of solute is transported into the fluid phase. As a result, the extraction yield is improved and characterized by positive sensitivities. Given that the results of the local sensitivity analysis depend on the operating point, the analysis was repeated at different pressures. The normalized sensitivity coefficients provide a basis for generalizing the discussion of the results and comparing the influence of various parameters. The normalized sensitivity results were found to vary within the range of -0.17 to 0.6. The overall effect of each parameter can be evaluated by integrating the sensitivity curves. Similar to the analysis of yield sensitivity, the integration was performed over extended simulation time (until all sensitivities approached zero). The values of cumulative integral are presented in Table \ref{table:Cumulative_integral}. The positive sign of these values indicates that increased pressure enhances the extraction yield under all investigated operating conditions. The order of magnitude of the integrals allows to identify at which operating condtions the pressure increase is the most influential.

>>>>

	One disadvantage of using local sensitivity coefficients is that they reflect absolute changes, which can be challenging to compare across different models or scenarios. To address this, normalized local sensitivity coefficients (defined as: $\tilde{S}(x;\Theta) = \frac{d G(x;\Theta)}{d\Theta} \frac{\Theta}{G(x;\Theta)}$) are introduced. They can be interpreted as the percentage change in the output per percentage change in the parameter. Their dimensionless nature makes the invariant under rescaling of model parameters.

@@

	%\subfile{Sections/Results_Sensitivity}
	\section{Results} \label{CH: Results}
	The details of the process model and parameters are provided in \citet{Sliczniuk2024}. The process model was validated in the following range of operating conditions: temperatures between $30 - 40 ^\circ C$, pressures between $100 - 200$ bar, and mass flow rates between $3.33-6.67 \cdot 10^{-5}$ kg/s. This study examines the impact of parameter change on the state space and extraction yield through local sensitivity analysis. The local sensitivity analysis results can be interpreted as the system's response to an infinitesimal deviation in parameters. {\color{blue}This work focus on two selected cases, but the same methodology can be extended to any other model parameter. The first case corresponds to the analysis of empirical correlations on the process model. The zero'th order term of the multiple linear regression for $\Upsilon$ (denoted as $\Upsilon(0)$) was selected. The second case corresponds to the investigation of the operating condition on the state-space. Particularly, on the influence of pressure variations on the model output. The first analysis is performed at the midpoint of the validated range: $35 ^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s. The second analysis is split into two steps. First, the influence of the pressure change on the state-variables (concentration of the solute in the sold and fluid phases) are considered at the midpoint of the validated range. Secondly, the yield analysis is conducted at the same temperature and mass flow-rate conditions but at different values of pressure (100, 125, 150, 175 and 200 bar).}

@@

	{\color{blue}The $\Upsilon$ term describes the curvature of decay function $\gamma$, and can be associated with uneven distribution of the solute in solid particles. As the zero'th order term is not directly affect by independent variables (as it represents the intercept), it physical interpretation becomes straight forward. The increment $\Upsilon(0)$ makes the $\gamma$ steeper, which means that the $D_i^R$ is multiplied by a smaller number, hence it is expected to observe slower extraction rate.

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CS_gamma.png}
		\caption{The effect of $\Upsilon(0)$ change on $c_s$}
		\label{fig:Sensitivty_Gamma_CS}
	\end{figure}

	Figure \ref{fig:Sensitivty_Gamma_CS} illustrates the sensitivity of the solute concentration in the solid phase to change into zero'th order term in the equation for $\Upsilon$. As discussed by \citet{Sliczniuk2024}, the system is far from the saturation point, which indicates that the concentration gradient is controlled mainly by $c_s$. Consequently, the extraction kinetic term act as independent on $c_f$, which explain a uniform response in solid phase visible in Figure \ref{fig:Sensitivty_Gamma_CS}. The positive value of $\frac{dc_s}{d\Upsilon(0)}$ can be interpreted as an slower loss of the solute from particles and is observed when the extraction slows down. Due to lower reaction rate more solute stays in the particles, but as the time approach infinity the solute will be washed out from the solid matrix. Consequently, it is expected to observed $\frac{dc_s}{d\Upsilon(0)}$ reaching a maximum and asymptotically diminish to zero.

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CF_gamma.png}
		\caption{The effect of $\Upsilon(0)$ change on $c_f$}
		\label{fig:Sensitivty_Gamma_CF}
	\end{figure}

	Figure \ref{fig:Sensitivty_Gamma_CF} illustrates the sensitivity of the solute concentration in the fluid phase with respect to change into zero'th order term in the equation for $\Upsilon$. As previously discussed, the extraction rate slowed down due to change of $\Upsilon(0)$ uniformly along the spatial domain. Consequently, the amount of delivered to the fluid phase is reduced by the same quantity along the spatial domain. That leads to decrease of the solute concentration $c_f$ in the fluid phase, which is characterized by negative sign of the $\frac{dc_f}{d\Upsilon(0)}$. The spatial distribution of $\frac{dc_f}{d\Upsilon(0)}$ in the figure is non-uniform, reflecting the effects of advection. Due to mass conservation, the remaining solute will reach the solid-fluid interface slower hence it is expected to observe a little positive values of $\frac{dc_f}{d\Upsilon(0)}$ in the later stage of extraction. Eventually it is expected to see sensitivities reaching zero asymptotically.

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/Y_gamma.png}
		\caption{The effect of $\Upsilon(0)$ change on $c_f$}
		\label{fig:Sensitivty_Gamma_Y}
	\end{figure}

	Figure \ref{fig:Sensitivty_Gamma_Y} show the normalized sensitivity curve of with respect to the yield. Due to slower extraction rate, the sensitivity curve is characterized by negative sign. The drop of the sensitivity curve can be interpreted as less solute coming out of the extractor if compared with the reference case. Then the sensitivity curve reach minimum, which can be also visible in Figure \ref{fig:Sensitivty_Gamma_CF}. Eventually, the remaining solute is released to the fluid phase and the sensitivity curve reach asymptotically zero. When the sensitivity function of a dynamic system approaches zero as time goes to infinity and remains at that value, it indicates that the system's long-term behavior becomes independent of the parameter in question. The parameter has only influence for the system's transient dynamics, it has no effect on the steady-state solution. This means that after the system has settled, variations in that parameter do not affect the final output. In terms of SFE, it means that all the solute has been removed from the system.}

@@

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CS_P.png}
		\caption{The effect of $P_{in}$ change on $c_s$}
		\label{fig:Sensitivty_P_CS}
	\end{figure}

@@

	\begin{figure}[!ht]
		\centering
		\includegraphics[trim = 0.0cm 0.0cm 0.0cm 0.0cm,clip,width=\columnwidth]{/Results_sensitivity/CF_P.png}
		\caption{The effect of $P_{in}$ change on $c_f$}
		\label{fig:Sensitivty_P_CF}
	\end{figure}

@@

@@

	Sensitivity analysis is a tool used to understand how the parameters of the model affect its output. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. This work implemented automatic differentiation to derive the sensitivity equations. The results of the local sensitivity analysis are model specific and consider only a small region of parameter space, so the conclusions derived from such an analysis are limited to local conditions. Although local sensitivity analysis can be performed with respect to any model parameter, this study focuses specifically on the effect of parameters of the empirical correlation and operating conditions on the state-space.

	{\color{blue}The motivation behind perform sensitivity analysis model validation and evaluation of parameters effect on the model output. By analysis the influence of parameters potential modelling errors and unfeasible behaviours can be identified. This is particularly useful technique, when introducing an empirical correlation. In this work, the influence of the intersect from the empirical correlation for $\Upsilon$ was tested. As expected, the analysis showed that increment of $\Upsilon(0)$ slows down the extraction rate, leading to negative values of $\frac{dc_s}{d\Upsilon(0)}$ and $\frac{dc_s}{d\Upsilon(0)}$, and alternating values of $\frac{dc_f}{d\Upsilon(0)}$. The physical interpretation of the $\Upsilon(0)$ match obtained results. The effect of the parameter change can be summarized by extreme and final values of the normalized sensitivity of the output (0.47 and 0, respectively).}

	pressure. Under the selected operating conditions, it was observed for all the cases that a pressure increment enhances mass transfer, leading to a faster mass transfer of solute from the particles. Consequently, negative sensitivities can be observed in the diagram for the solid phase. The corresponding response in the fluid phase is characterized by positive-valued sensitivities, indicating that a greater amount of solute is transported into the fluid phase. As a result, the extraction yield is improved and characterized by positive sensitivities. Given that the results of the local sensitivity analysis depend on the operating point, the analysis was repeated at different pressures. The normalized sensitivity coefficients provide a basis for generalizing the discussion of the results and comparing the influence of various parameters. The normalized sensitivity results were found to vary within the range of -0.17 to 0.6. The overall effect of each parameter can be evaluated by integrating the sensitivity curves. Similar to the analysis of yield sensitivity, the integration was performed over extended simulation time (until all sensitivities approached zero). The values of cumulative integral are presented in Table \ref{table:Cumulative_integral}. The positive sign of these values indicates that increased pressure enhances the extraction yield under all investigated operating conditions. The order of magnitude of the integrals allows to identify at which operating condtions the pressure increase is the most influential.
